---
title: "Munge"
author: "Andy Choens"
date: "Week 03"
output: 
  html_notebook:
    toc: true
    toc_float:
      toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: cerulean
---



# Tonight

- Data Life Cycle
- Last Week Tonight
    - Some `mpg` talk
    - A ggplot review
- Intro To Data Munging



# Data Life Cycle

```{r echo=false, fig.align='center'}
url <- "https://d33wubrfki0l68.cloudfront.net/571b056757d68e6df81a3e3853f54d3c76ad6efc/32d37/diagrams/data-science.png"
knitr::include_graphics(url)
```

- Last week, an introduction to data viz.
- This week, an introduction data munging (Tidy/Transform) & some data viz.
- Next week, more data viz and more data munging.

**BIG IDEA:** The data you have is ALMOST NEVER the data you want.


# Last Week Tonight

```{r}
library(tidyverse)
data(mpg)
mpg
```

- The grain of the `mpg` data is car model.
    - Each row is a different model.
    - Brands span many rows.
    - Years span many rows.
    - We have data about the car model, but not about individual cars.
- So what?
    - This data tells us something about the mileage of car models for sale.
    - It tells us nothing about the mileage of cars actually sold.
    - It tells us nothing, really, of mileage of cars on the road.
    
For example:

```{r}
mpg %>% filter(manufacturer == "chevrolet")
```
We cannot compare the number of Suburbans sold to the number of Malibus sold.

But, we can learn about this data set itself.

```{r}
ggplot(data = mpg, aes(x = class)) + 
   geom_bar(aes(y = 100*(..count..)/sum(..count..)), fill = "darkred") +
   ylim(0, 100)
```
A bit more complicated than what your lab, but you should recognize it by now. 
There are three parts:

1. `ggplot(data = mpg, aes(x = class)) + ` DEFINES THE DATA
2. `geom_bar(aes(y = 100*(..count..)/sum(..count..)), fill = "darkred") +` DEFINES WHAT
3. `ylim(0, 100)` EXTRA JAZZ

This plot shows PERCENTS, not COUNTS. Thus, it is appropriate for it to show the
full range of possible percents (0 - 100). Be very careful looking at a graph
showing percents that doesn't show you where 0 and 100 are relative to the data
shown. If they don't show you that, assume you are being manipulated (or the
analyst is a amateur).



# Some 2020 DAta

OK, that data set is cute, but can we get more data, newer data?

YES WE CAN! [https://www.fueleconomy.gov/](https://www.fueleconomy.gov/)

The US Federal Government publishes annual estimates for the mileage of 
cars sold in the US each year.

- You do not need to download this data. I did that for you and I added it here.
- You may, if you are interested, look at the [script I wrote](./data/get_mpg.R).
- If you want to look yourself, at the bottom of the page, "Download EPA's MPG Ratings"

## Column Headers

If you looked at the script, you know I simplified the column names to make them
easier to use in R. However, my column names necessarily lose some detail. Therefore,
to preserve the original full detail:

```{r}
## This tibble show us my shortened column name next to the original full
## column name. Don't worry aout this code. I'm providing this to make your
## life easier.
tibble(ColName = names(mpg2020), Original = attributes(mpg2020)$label)
```

Above, we drew a bar plot of car class. The equivalent plot, with this data set
would be:

```{r}
## Differences:
## - I rotated the graph to make the class names easier to read.
## - The column name used is different.
## - There are more classes in this data set, so that's neat.
ggplot(data = mpg2020, aes(y = CarlineClassDesc)) + 
   geom_bar(aes(x = 100*(..count..)/sum(..count..)), fill = "darkred") +
   xlim(0, 100)
```

This data set has everything in the original mpg data, but it is newer and, 
well, more. But, how can we:

- Draw this plot, for Chevrolets only?
- Calculate the mileage, for Standard Pick-ups?
- Figure out which manufacturer has the most models for sale?
- Figure out who sold the most cars. (Oops, you can't do that with this data!)

The data needed to answer these AND MANY OTHER questions are here, but we cannot
easily do this without first transforming (munging) the data. And this is a real
data set and you can imagine real questions that could be answered with it, but
to do so, we must be able to transform the data


# DATA MUNGING

Let me encourage you to read/watch:

- [R4DS: Data Transformation](http://r4ds.had.co.nz/transform.html)
- [dplyr intro | Data manipulation in R](https://www.youtube.com/watch?v=rm0BQSWoJlc)

We transform data using a series of verbs from the dplyr package (part of the 
Tidyverse). Each very performs a specific data transformation step.

- Filter: Only keep the rows of data that you need for the analysis. For 
  example, we could use this function to drop all rows not made by Chevrolet.
    - Function: `filter()`
    - ==
    - <, <=, >, >=
    - is.na()
    - !
- Get Summary Stats: Provide summary statistics such as the average value of a
  column, or the number of lines in the data set.
    - Function: `summarize()`
    - min/max
    - mean/sd
    - n
- Group: Group, or stratify our results by a column. This is similar to
  `facet_wrap()` we learned about last week, but more flexible.
    - Function: `group_by()`
- Sort: Sort, or arrange, the results by the values in column.
    - Function: `arrange()`
    - By default, returns are sorted in ascending order.
    - `desc()` Returns can be sorted in descending order.

```{r}
url <- "https://upload.wikimedia.org/wikipedia/en/b/bd/ThePlaneteers.JPG"
knitr::include_graphics(url)
```

With their powers combined, we can perform many interestng transformations of
our data, to answer questions the original data cannot answer.

We will learn about other data transofrmations in the coming weeks. This is
enough to get us started doing something interesting.

## Filter

Uses Boolean logic to filter our data.

- ==
- <, <=, >, >=
- is.na()
- !

```{r}
chevies <- filter(mpg2020, Division == "Chevrolet")
chevies
```

In case you didn't realize this, General Motors owns/manufacturers Chevrolet 
and several car brands such as Cadillac & GMC.

```{r}
paste("Number Rows mpg2020:", nrow(mpg2020))
paste("Number Rows chevies:", nrow(chevies))
```

We can look at cars with 4 cylinders.

```{r}
four_cylinders <- filter(mpg2020, NCyl == 4)
four_cylinders
```

Or cars with less than 6 cylinders!

```{r}
lt_six_cylinders <- filter(mpg2020, NCyl < 6)
lt_six_cylinders
```

In case you didn't realize this, there many different kinds of Honda Civic. We
can find them all without having to know the Carline name for each!

```{r}
civics <- filter(mpg2020, str_detect(Carline, "CIVIC"))
civics
```

And, this isn't just academic. Perhaps I want to see a density plot of Honda
Civic city gas mileage.

```{r}
ggplot(civics, aes(x = City)) + geom_density(color = "darkgreen") 
```

This differs from the over-all dramatically!

```{r}
ggplot(mpg2020, aes(x = City)) + geom_density(color = "darkred") 
```

Some take-aways:

1. How do you get a Honda Civic that can only get 22/23 mpg in the city?
2. PLEASE observer how these two graphs peak in very different places!
3. The first parameter in the filter command AND the ggplot command is data.

## Summarize

- Function: `summarize()`
- min/max
- mean/sd
- n

A summary statistic, hence summarize, is a single number that tells us something
about the data in that column.

```{r}
summarize(civics,
          N = n(),
          CityAvg = mean(City),
          CitySD = sd(City),
          CityMin = min(City),
          CityMax = max(City))
```

Does this make sense, in the context of our data set and density plot above?

## Group By

Stratify the results of a summarize by another variable. Every year, this gives
students a hard time. So here's an example I hope helps.

- Function: `group_by()`

```{r}
grouped_civics <- group_by(civics, Carline)
summarize(grouped_civics,
          N = n(),
          CityAvg = mean(City),
          CitySD = sd(City),
          CityMin = min(City),
          CityMax = max(City))
```

Wait, seriously? There's a five-door Civic?

- The function `group_by` creates a grouped data set which is nearly 
  indistinguishable from an un-grouped data set.
- This command is not useful on it's own.
- It is very useful when combined with summarize because a grouped data set
  summarizes differently than an un-grouped data set.
- Again, notice how the data set is ALWAYS the first parameter.
    - There must be some reason I keep harping on this. . . . 
  
```{r}
## The only differentiating feature is the part that says Groups: Carline [3]
grouped_civics
```
  
As you can probably imagine, if you do this a bunch, you wind up with a bunch of
awkwardly named data sets. I've got six or seven going now, and I'm just messing
around.

Fortunately, there is a better way! Pipes!

```{r}
civics %>%
   group_by(Carline) %>%
   summarize(
          N = n(),
          CityAvg = mean(City),
          CitySD = sd(City),
          CityMin = min(City),
          CityMax = max(City))
```

This is EXACTLY the same as what we did above, but we did it in a single step.
The pipe symbol `%>%` "pipes" the output from one function to the input of the
next function and works when the first parameter of the function is the data.

That's why I kept pointing this out. We can do all sorts of piping!

```{r}
easy_summary_civics <- 
   mpg2020 %>%
   filter(str_detect(Carline, "CIVIC")) %>%
   group_by(Carline) %>%
   summarize(
          N = n(),
          CityAvg = mean(City),
          CitySD = sd(City),
          CityMin = min(City),
          CityMax = max(City))
easy_summary_civics
```

Thanks to the pip operator, we can avoid creating a million data-step data sets.
This is kinda-sorta available in SAS, but it is not taught as a best practice.
In modern R programming, using the pipe is considered best practice and is how
this stuff gets taught.

## Arrange

After performing one or more data transformations, R will return our data to us
in alphabetical order, based on the first column or so. For example, see above.

But, that isn't ALWAYS how we want it.

- Function: `arrange()`
- By default, returns are sorted in ascending order.
- `desc()` Returns can be sorted in descending order.

```{r}
easy_summary_civics %>% arrange(CityAvg)
easy_summary_civics %>% arrange(desc(CityAvg))
```

This function is a useful function on your labs!!!!!! For example, say I ask a
question like this:

> Task X: Which Car in mpg2020 has the HIGHEST highway gas mileage?

So, there are over 1,000 rows in mpg2020. I don't think you want to go searching
through that by hand. In fact, I KNOW you don't. (If you do, let's talk after 
class.)

Fortunately, arrange makes it easy.

```{r}
mpg2020 %>% arrange(desc(Hwy)) %>% slice(5)
```


# Using Data Responsibly 

Always look for documentation to help you understand the: 

- content of the data (columns)
- sampling/inclusion/exclusion criteria (as relevant/appropriate)
- how much data is there (this goes back to 1978)
- caveats, changes in the data over time.
- Any aggregations we can use to validate our analyses.

For example, this data does NOT include EVERY vehicle sold in the US. From the
PDF documentation on the EPA website, we can see that some vehicles are exempt.

Light-duty fuel economy regulations do not apply to:

- Sport utility vehicles (SUVs) and passenger vans with a gross vehicle weight
  rating (GVWR) of more than 10,000 pounds—GVWR is the vehicle weight plus
  carrying capacity
- Other vehicles with a GVWR of 8,500 pounds or more or a curb weight over
  6,000 pounds. Therefore, manufacturers do not have to estimate their fuel 
  economy, and fuel economy labels are not posted on their windows. Also, fuel 
  economy information on some vehicles was not available in time to be included 
  in this guide. More up-to-date information can be found at fueleconomy.gov.

This exclusion is neither "right" nor "wrong", but it does bias the data. There
are reasons the EPA does this. You may or may not agree with these reasons. My 
point is simple - you need to be aware of these kinds of exclusions and biases.
If you don't look for this information, you are using the data irresponsibly.


https://www.caranddriver.com/features/a15388892/the-truth-about-epa-city-highway-mpg-estimates/